Data-based completion of grey-box models with neural networks

### 2020-11-23 systemID_NNs.tex
add scripts.

@@

\section{Process modeling}
Consider a nonlinear plant evolving in continuous time as
\begin{align*}
  \dot x &= f_p(x, u) \\
  y &= h_p(x)
\end{align*}
in which $x$ is the plant state (vector), $u$ is the manipulated control input,
and $y$ is the measurement. Assume that we don't have any 
disturbances/measurement and process noise 
for now, and we'll add those after we 
understand the issues in the nominal case.

@@

in which $x_g$ are the grey-box states, which are some subset 
of the plant states (denote $x_g = Gx$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k+1-N_p:k} &= [y(k)', y(k-\Delta)', ..., y(k-(N_p-1)\Delta)']' \\
  \mathbf{u}_{k+1-N_p:k-1} &= [u(k-\Delta)', u(k-2\Delta)', 
                               ..., u(k-(N_p-1)\Delta)']'
\end{align*}

in which $N_p$ is the number of 
past measurements and
$\Delta$ is the sample time between two consecutive
measurements. There exists a $\phi(\cdot)$ 
and a number of delay 
time steps $N_p$, such that the plant state
at a given time can be written as
$x(k) = \phi(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})$.
For linear plants, this function $\phi(\cdot)$
is known analytically in terms of the plant dynamic
matrices.

We augment the grey-box model using 
neural networks as follows -
\begin{align*}
\dot x_g &= f_g(x_g, u) + 
f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k})\\
y &= h_g(x_g) + h_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})
\end{align*}
in which $f_{NN}(\cdot)$ and $h_{NN}(\cdot)$
are standard feedforward neural networks. We add the current 
control $u(k)$ as an input to the network $f_{NN}(\cdot)$ because 
there may be a mismatch in the actuator to the grey-box dynamics
as well.

It is expected that the functions $f_{NN}(\cdot)$ 
and $h_{NN}(\cdot)$ approximate
the following differences between the true grey-box state dynamics
and true measurement function with what is modeled.
\begin{align*}
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k}) &\approx 
      Gf_p(x, u) - f_g(Gx, u) \\
      h_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) &\approx
      h_p(x) - h_g(Gx) 
\end{align*}  
in which $x(k) = \phi(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})$
is an unknown function of past measurements and inputs.

\subsection{Conversion to discrete time and training}
We note that the network $f_{NN}(\cdot)$ is an addition 
to the grey-box dynamics in continuous time, whereas 
$h_{NN}(\cdot)$ is an addition directly to the 
measurement function. The Runge-Kutta-4 (RK4) method 
is used for the discretization as follows

\begin{align*}
  k_1 &= f_g(x_g(k), u(k)) + 
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k}) \\
  k_2 &= f_g(x_g(k) + (\Delta/2)k_1, u(k)) + 
  f_{NN}(\mathbf{y}_{k+1.5-N_p:k+0.5}, \mathbf{u}_{k+1.5-N_p:k+0.5})\\
  k_3 &= f_g(x_g(k) + (\Delta/2)k_2, u(k)) + 
  f_{NN}(\mathbf{y}_{k+1.5-N_p:k+0.5}, \mathbf{u}_{k+1.5-N_p:k+0.5})\\
  k_4 &= f_g(x_g(k) + \Delta k_3, u(k)) + 
  f_{NN}(\mathbf{y}_{k+2-N_p:k+1}, \mathbf{u}_{k+2-N_p:k+1})\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) + 
  h_{NN}(\mathbf{y}_{k+2-N_p:k+1}, \mathbf{u}_{k+2-N_p:k+1})
\end{align*}
in which the sequences $\mathbf{y}_{k+1.5-N_p:k+0.5}$ 
and $\mathbf{u}_{k+1.5-N_p:k+0.5}$ are defined as 
\begin{align*}
  \mathbf{y}_{k+1.5-N_p:k+0.5} &= 
    [y(k + \Delta/2)', y(k-\Delta/2)', ..., y(k-(N_p-1)\Delta + \Delta/2)']' \\
    \mathbf{u}_{k+1.5-N_p:k+0.5} &= 
  [u(k + \Delta/2)', u(k-\Delta/2)', ..., u(k-(N_p-1)\Delta + \Delta/2)']' 
\end{align*}
We assume a zero order hold on the control inputs between two consecutive 
MPC execution instants, such that $u(k + \Delta/2) = u(k)$. To get the 
measurements between two sample instants (for e.g., $y(k + \Delta/2)$) for 
the RK4 predictions, linear interpolation is used.

The following optimization problem is solved to compute 
the weights in the neural networks 
\begin{align*}
 \underset{W_{fi}, b_{fi}, W_{hi}, b_{hi}}{\textnormal{min}} \sum_{k=0}^{N_s-1} 
 \dfrac{1}{2}(\hat{y}(k) - y(k))^2
\end{align*}
in which ($W_{fi}, b_{fi}, W_{hi}, b_{hi}$) are the weights in the 
two neural networks ($f_{NN}(\cdot)$ and $h_{NN}(\cdot)$), 
$\hat{y}$ are the predictions of the augmented
grey-box + neural network model, and $y$
are the measurements from the plant.
Note that this optimization requires an estimate 
of the initial grey-box state. We assume that 
before the open-loop system identification experiment 
performed for data collection, 
the plant is being operated at some 
steady-state such that input to output
responses of the plant and grey-box model are aligned with 
the use of an integrating disturbance model. We can use 
the grey-box state at that steady state as the initial condition 
for the above multi-step-ahead prediction-error
optimization.

@@

\textbf{Plant --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
These ODEs represent the dynamics of the reaction system
$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
height and temperature. The control input is $u = C_{A0}$,
the states are $x = [C_A, C_B, C_C]'$, 
and the measurements are $y = [C_A, C_B]'$. 
Assume the following 
parameters for this model, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.3 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.2 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
The plant model 
is linear ($dx/dt = Ax + Bu$) and 
observable with the choices of 
the parameters and measurements.

\textbf{Model --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
\end{align*}
i.e, we assume that we don't know about the second reaction while 
performing grey-box modeling.

The current state $x(k)$ can be constructed using the history of 
past measurements and inputs as 
\begin{equation*}
  x(k) = A^{N_p-1}_d
  \mathcal{O}^{\dagger}(\mathbf{y}_{k+1-N_p:k} - 
                        \mathcal{B}\mathbf{u}_{k+1-N_p:k-1}) + \\ 
      \mathcal{B}_{nN_p-n+1:nN_p}\mathbf{u}_{k+1-N_p:k-1}
\end{equation*}

@@

Since the measurements are a simple linear function 
of the grey-box and plant states, we use only one 
neural network to 
augment the grey-box dynamics. 
The expected function 
to be approximated with the network is
\begin{align*}
  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) \approx 
  \begin{bmatrix}
  0 & 0 & 0 \\  
  0 & -k_2 & k_3 
  \end{bmatrix}x(k)
\end{align*}

@@

@@

We notice that the augmented model accurately predicts the plant 
behavior. The steady cost profile (Cost = $c_1C_{A0} - c_2C_B$)
is shown in Figure \ref{fig:cost_linear}.

\begin{figure}[!h]
  \centering
  \includegraphics[page=1, width=\textwidth]{tworeac_ssopt_lin.pdf}
  \caption{Steady state cost curve for 
           the linear plant, grey-box, and hybrid model.}
  \label{fig:cost_linear}
\end{figure}

\subsection{Nonlinear plant}
The plant is made nonlinear by changing the side reaction 
to a third order reaction ($A \rightarrow B $ and  $3B \rightleftharpoons C$),
and is now simulated with the following ODEs
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{A0} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

The grey-box model is kept the same. The parameters 
in the model are modified to 
$k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.01 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.05 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
An augmented model with a neural
network architecture [8, 4, 2] is trained on 360 samples (6 hours). 
The tanh activation is used in the network. The figure below shows 
validation performances of the grey-box and augmented 
hybrid models.

@@

We notice that the augmented model accurately predicts the plant 
behavior. The steady cost profile (Cost = $c_1C_{A0} - c_2C_B$)
is shown in Figure \ref{fig:cost_nonlinear}. Upon optimization 
using casadi, the optimum cost and control for the plant are: $-19.62 \$$ 
and 1.13 $\textnormal{mol}/\textnormal{m}^3$, and for the hybrid 
grey-box and NN model are: $-19.91 \$$ 
and 1.11 $\textnormal{mol}/\textnormal{m}^3$.

\begin{figure}[!h]
  \centering
  \includegraphics[page=1, width=\textwidth]{tworeac_ssopt_nonlin.pdf}
  \caption{Steady state cost curve for the nonlinear plant, grey-box, and hybrid model.}
  \label{fig:cost_nonlinear}
\end{figure}

>>>>

\newcommand{\useq}{\mathbf{u}}
\newcommand{\xseq}{\mathbf{x}}
\newcommand{\bbR}{\mathbb{R}}
\newcommand{\bbW}{\mathbb{W}}
\newcommand{\bbU}{\mathbb{U}}
\newcommand{\bbI}{\mathbb{I}}
\newcommand{\bbX}{\mathbb{X}}
\newcommand{\bbP}{\mathbb{P}}

@@

\section{Process modeling}
Consider a nonlinear plant evolving in continuous time as
\begin{align*}
  \dot x_p &= f_p(x_p, u, w) \\
  y &= h_p(x_p) + v
\end{align*}
in which $x_p \in \bbR^{n_p}$ is the plant state, 
$u \in \bbU$ is the manipulated control input,
and $y \in \bbR^p$ is the measurement.

@@

in which $x_g \in \bbR^{n_g}$ are the grey-box states, 
which are some subset 
of the plant states (denote $x_g = Gx_p$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k-N_p:k-1} &= [y(k-\Delta)', ..., y(k-N_p\Delta)']' \\
  \mathbf{u}_{k-N_p:k-1} &= [u(k-\Delta)', 
                               ..., u(k-N_p\Delta)']'
\end{align*}

in which $N_p$ is the number of 
past measurements and
$\Delta$ is the sample time between two consecutive
measurements. There exists a $\phi(\cdot)$ 
and a number of delay 
time steps $N_p$, such that the plant state
at a given time can be written as
$x_p(k) = \phi(\mathbf{y}_{k-N_p:k}, \mathbf{u}_{k-N_p:k-1})$.
For linear plants, this function $\phi(\cdot)$
is known analytically in terms of the plant dynamic
matrices.

We augment the grey-box model using 
neural networks as follows -
\begin{align*}
\dot x_g &= f_g(x_g, u) + 
f_N(h_g(x_g), z, u)\\
y &= h_g(x_g) 
\end{align*}
in which $f_N(\cdot)$ is a feedforward neural network
and $z(k) = [\mathbf{y}'_{k-N_p:k-1}, \mathbf{u}'_{k-N_p:k-1}]'$.

Alternative modeling choices are:
\begin{itemize}
	\item Black-box: $y^+ = f_N(y, z, u)$, (purely black-box)
	\item Hybrid 1: $\dot x_g = f_N(x_g, z, u), \ y = h_g(x_g)$, 
  (states have meaning)
  \item Hybrid 2: $\dot x_g = f_g(x_g, u), \ y = h_g(x_g) + h_N(z)$, 
	(model residual)
	\item Hybrid 3: $\dot x_g = f_g(x_g, u, p(h_g(x_g), z)), \ y = h_g(x_g)$, (use NNs to learn some parameters)
\end{itemize}

%It is expected that the functions $f_{NN}(\cdot)$ 
%and $h_{NN}(\cdot)$ approximate
%the following differences between the true grey-box state dynamics
%and true measurement function with what is modeled.
%\begin{align*}
%  f_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k}) &\approx 
%      Gf_p(x, u) - f_g(Gx, u) \\
%      h_{NN}(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) &\approx
%      h_p(x) - h_g(Gx) 
%\end{align*}  
%in which $x(k) = \phi(\mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k-1})$
%is an unknown function of past measurements and inputs.

\subsection{Conversion to discrete time and training}
We note that the network $f_N(\cdot)$ is an addition 
to the grey-box dynamics in continuous time. 
The Runge-Kutta-4 (RK4) method 
is used for the discretization as follows

\begin{align*}
  k_1 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k-N_p:k-1}, \mathbf{u}_{k+1-N_p:k}) \\
  x_g & := x_g + (\Delta/2)k_1 \\
  k_2 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k+1-N_p:k})\\
  x_g & := x_g + (\Delta/2)k_2 \\
  k_3 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k+1-N_p:k})\\
  x_g & := x_g + \Delta k_3 \\
  k_4 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k})\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) 
\end{align*}
in which the sequence $\mathbf{y}_{k+0.5-N_p:k-0.5}$ is obtained by 
linear interpolation.

The following optimization problem is solved to compute 
the weights in the neural network
\begin{align*}
\underset{W_i, b_i}{\textnormal{min}} \sum_{k=0}^{N_s-1} 
\dfrac{1}{2}(\hat{y}(k) - y(k))^2
 \end{align*}
in which $W_i$, $b_i$ are the weights in each hidden layer of the NN, 
$N_s$ is the number of training samples, 
$\hat{y}$ are the predictions of the augmented
grey-box + neural network model, and $y$
are the measurements from the plant.
Note that this optimization requires an estimate 
of the initial grey-box state. We assume that 
before the open-loop system identification experiment 
performed for data collection, 
the plant is being operated at some 
steady-state such that input to output
responses of the plant and grey-box model are aligned with 
the use of an integrating disturbance model. We can use 
the grey-box state at that steady state as the initial condition 
for the above multi-step-ahead prediction-error
optimization.

@@

\textbf{Plant --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
These ODEs represent the dynamics of the reaction system
$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
height and temperature. The control input is $u = C_{Af}$,
the states are $x = [C_A, C_B, C_C]'$, 
and the measurements are $y = [C_A, C_B]'$. 
Assume the following 
parameters for this model, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.3 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.2 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
The plant model 
is linear ($dx/dt = Ax + Bu$) and 
observable with the choices of 
the parameters and measurements.

\textbf{Model --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
\end{align*}
i.e, we assume that we don't know about the second reaction while 
performing grey-box modeling.

The current plant state $x_p(k)$ can be constructed using the history of 
past measurements and inputs as 
\begin{equation*}
  x_p(k) = A^{N_p-1}_d
  \mathcal{O}^{\dagger}(\mathbf{y}_{k+1-N_p:k} - 
                        \mathcal{B}\mathbf{u}_{k+1-N_p:k-1}) + \\ 
      \mathcal{B}_{nN_p-n+1:nN_p}\mathbf{u}_{k+1-N_p:k-1}
\end{equation*}

@@

Since the measurements are a simple linear function 
of the grey-box and plant states, we use only one 
neural network to 
augment the grey-box dynamics. 
The expected function 
to be approximated with the network is
\begin{align*}
  f_N(\mathbf{y}_{k-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) \approx 
  \begin{bmatrix}
  0 & 0 & 0 \\  
  0 & -k_2 & k_3 
  \end{bmatrix}x_p(k)
\end{align*}

@@

THe Figures \ref{fig:cost_nonlinear_5hours} and 
\ref{fig:cost_nonlinear_5hours}

@@

We notice that the augmented model accurately predicts the plant 
behavior. The steady cost profile (Cost = $c_1C_{Af} - c_2C_B$)
is shown in Figure \ref{fig:cost_linear}.

\begin{figure}[!h]
  \centering
  \includegraphics[page=3, width=\textwidth]{tworeac_plots_lin.pdf}
  \caption{Steady state cost curve for 
           the linear plant, grey-box, and hybrid model.}
  \label{fig:cost_linear}
\end{figure}

\subsection{Nonlinear plant}
The plant is made nonlinear by changing the side reaction 
to a third order reaction ($A \rightarrow B $ and  $3B \rightleftharpoons C$),
and is now simulated with the following ODEs
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

The grey-box model is kept the same. The parameters 
in the model are modified to 
$k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.01 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.05 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.

\newpage
Three models are compared: black-box, residual, and hybrid model with additive 
NN to the grey-box dynamics. The chosen model architectures 
are [9, 16, 16, 2], [9, 16, 16, 2], and [8, 16, 2] respectively.
The Figure \ref{fig:validation_nonlinear} shows model validation on open loop 
data after training with 10 hours of data.

The Figures \ref{fig:cost_nonlinear_5hours} and \ref{fig:cost_nonlinear_10hours}
show the steady state cost curve after training with 5 and 10 hours 
of data respectively. We notice that the hybrid model is more data efficient 
compared to the alternatives.

@@

%We notice that the augmented model accurately predicts the plant 
%behavior. The steady cost profile (Cost = $c_1C_{A0} - c_2C_B$)
%is shown in Figure \ref{fig:cost_nonlinear}. Upon optimization 
%using casadi, the optimum cost and control for the plant are: $-19.62 \$$ 
%and 1.13 $\textnormal{mol}/\textnormal{m}^3$, and for the hybrid 
%grey-box and NN model are: $-19.91 \$$ 
%and 1.11 $\textnormal{mol}/\textnormal{m}^3$.

\begin{figure}[!h]
  \centering
  \includegraphics[page=3, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Steady state cost curve for the nonlinear plant, grey-box, 
  black-box, residual, and hybrid model after training with 5 hours of data.}
  \label{fig:cost_nonlinear_5hours}
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[page=4, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Steady state cost curve for the nonlinear plant, grey-box, 
  black-box, residual, and hybrid model after training with 10 hours of data.}
  \label{fig:cost_nonlinear_10hours}
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[page=1, width=\textwidth]{tworeac_distmodels_nonlin.pdf}
  \caption{Steady state grey-box states and disturbance estimates based on 
  output and input disturbance models}
  \label{fig:cost_nonlinear_10hours}
\end{figure}
